---
title: "EDA of Water Table Data"
author: "**Istiak Ahmed***, SO, ASICT Division, BARI*"
output: html_notebook
---

```{r}
rm(list = ls())
```

```{r}
# Load required libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
library(plotly)
library(ggridges)
library(viridis)
library(ggExtra)
library(patchwork)
library(zoo)
library(tsibble)
library(feasts)
library(imputeTS)
library(gridExtra)
library(corrplot)
library(ggmap)  # For mapping (optional)
library(leaflet)  # For interactive mapping
library(colorspace)
```

# ============================
# 1. DATA LOADING AND CLEANING
# ============================

```{r}
# Read the data
df <- read.csv("Water_Table_Data.csv", stringsAsFactors = FALSE)
```

```{r}
# Check structure
str(df)
head(df)
```

```{r}
# Convert Date to proper date format
df$Date <- dmy(df$Date)  # Using lubridate's dmy function

# Check for any parsing issues
summary(df$Date)
```

```{r}
# Check for duplicate rows
duplicate_count <- sum(duplicated(df))
cat("Number of duplicate rows:", duplicate_count, "\n")
```

```{r}
# Sort by District and Date
df <- df %>%
  arrange(District, Date)
```

```{r}
# Check for outliers (extreme values)
summary(df$Water_Table_m)
boxplot(df$Water_Table_m, main = "Boxplot of Water Table Depth",
        ylab = "Depth (m)")
```

```{r}
# Identify potential data entry errors (e.g., 36.6 for Tangail)
potential_errors <- df %>%
  filter(Water_Table_m > 12.23 | Water_Table_m < 0) %>%
  arrange(desc(Water_Table_m))
```

```{r}
as.data.frame(table(potential_errors$District))
```

```{r}
potential_errors %>%
  group_by(District) %>%
  summarise(
    count = n(),
    across(Water_Table_m, list(min = min, mean = mean, max = max), .names = "{.fn}")
  )
```


# ========================================
# 2. DATA COVERAGE AND SUMMARY STATISTICS
# ========================================

```{r}
# Date range
date_range <- range(df$Date, na.rm = TRUE)
cat("Data covers from", format(date_range[1], "%Y-%m-%d"), 
    "to", format(date_range[2], "%Y-%m-%d"), "\n")
```

```{r}
# Number of unique districts
n_districts <- length(unique(df$District))
cat("Number of unique districts:", n_districts, "\n")
```

```{r}
# Observations per district
district_counts <- df %>%
  group_by(District) %>%
  summarise(n_obs = n()) %>%
  arrange(desc(n_obs))

district_counts
```


```{r}
# Read the well count data
well_counts <- read.csv("Well_count.csv", stringsAsFactors = FALSE)
```

```{r}
# Clean the well count data
well_counts_clean <- well_counts %>%
  # Remove rows with missing District
  filter(!is.na(District) & District != "") %>%
  # Group by District to get well count and total observations
  group_by(District) %>%
  summarise(
    Well_Count = n_distinct(Well_ID[!is.na(Well_ID) & Well_ID != ""]),
    Total_Observations = sum(Observation, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r,fig.height=8,fig.width=10}
enhanced_obs_plot <- well_counts_clean %>%
  mutate(District = reorder(District, Total_Observations)) %>%
  ggplot(aes(x = District, y = Total_Observations)) +
  geom_bar(stat = "identity", aes(fill = Well_Count), alpha = 0.8) +
  scale_fill_viridis_c(name = "Number of Wells", direction = -1) +
  # Add text labels for well count
  geom_text(aes(label = paste("Wells:", Well_Count), 
                y = Total_Observations + max(Total_Observations) * 0.02),
            size = 2.5, hjust = 0) +
  coord_flip() +
  labs(title = "Observations per District with Well Count Annotation",
       subtitle = "Bar color indicates number of wells | Text shows exact well count",
       x = "District", y = "Number of Observations") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 7),
        legend.position = "none")

print(enhanced_obs_plot)
```

```{r}
# Summary statistics by district
summary_stats <- df %>%
  group_by(District) %>%
  summarise(
    mean_depth = mean(Water_Table_m, na.rm = TRUE),
    median_depth = median(Water_Table_m, na.rm = TRUE),
    sd_depth = sd(Water_Table_m, na.rm = TRUE),
    min_depth = min(Water_Table_m, na.rm = TRUE),
    max_depth = max(Water_Table_m, na.rm = TRUE),
    n_obs = n()
  ) %>%
  arrange(desc(mean_depth))

summary_stats
```


# ===========================
# 3. TEMPORAL VISUALIZATIONS
# ===========================

## 3.1 Overall "Spaghetti" Plot (all districts)

```{r,fig.width=10}
# Note: This can be very busy, so we'll use transparency
spaghetti_plot <- ggplot(df, aes(x = Date, y = Water_Table_m, 
                                  color = District, group = District)) +
  geom_line(alpha = 0.4) +
  labs(title = "Water Table Depth Over Time - All Districts",
       x = "Date", y = "Water Table Depth (m)",
       color = "District") +
  theme_classic()+
  theme(legend.position = "nonr")  # Hide legend to avoid clutter

print(spaghetti_plot)
```

## 3.2 Time Series for a selection of districts (based on different regions)

```{r,fig.width=10}
selected_districts <- c("Rangpur", "Barisal", "Chattogram", "Mymensingh", 
                        "Dhaka", "Rajshahi", "Sylhet", "Khulna")

selected_plot <- df %>%
  filter(District %in% selected_districts) %>%
  ggplot(aes(x = Date, y = Water_Table_m, color = District)) +
  geom_line(size = 0.8) +
  facet_wrap(~District, scales = "free_y", ncol = 2) +
  labs(title = "Water Table Trends - Selected Districts",
       x = "Date", y = "Water Table Depth (m)") +
  theme(legend.position = "none")

print(selected_plot)
```

# ========================
# 4. SEASONALITY ANALYSIS
# ========================

```{r}
# Extract year and month
df <- df %>%
  mutate(
    Year = year(Date),
    Month = month(Date, label = TRUE, abbr = TRUE),
    Year_Month = floor_date(Date, "month")
  )
```

## 4.1 Monthly averages across all districts

```{r}
monthly_avg <- df %>%
  group_by(Month) %>%
  summarise(
    mean_depth = mean(Water_Table_m, na.rm = TRUE),
    sd_depth = sd(Water_Table_m, na.rm = TRUE)
  )

seasonal_plot <- ggplot(monthly_avg, aes(x = Month, y = mean_depth, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(ymin = mean_depth - sd_depth, 
                    ymax = mean_depth + sd_depth), 
                width = 0.2, alpha = 0.5) +
  labs(title = "Seasonal Pattern - Monthly Average Water Table Depth",
       x = "Month", y = "Water Table Depth (m)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(seasonal_plot)
```

## 4.2 Seasonal heatmap (Month x Year)

```{r}
seasonal_heatmap_data <- df %>%
  group_by(Year, Month) %>%
  summarise(mean_depth = mean(Water_Table_m, na.rm = TRUE)) %>%
  filter(Year >= 2015 & Year <= 2024)  # Focus on complete years

heatmap_plot <- ggplot(seasonal_heatmap_data, 
                       aes(x = Month, y = Year, fill = mean_depth)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Depth (m)", direction = -1) +
  labs(title = "Seasonal Heatmap: Water Table Depth",
       x = "Month", y = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(heatmap_plot)
```


## 4.3 Ridgeline plot to show seasonal distribution

```{r,fig.width=10}
ridge_plot <- ggplot(df, aes(x = Water_Table_m, y = Month, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Depth (m)", direction = -1) +
  labs(title = "Distribution of Water Table Depth by Month",
       x = "Water Table Depth (m)", y = "Month")+
  theme_classic()

print(ridge_plot)
```


## 4.4 Seasonal decomposition for a specific district

```{r}
# Choose a district with good data coverage
decomp_district <- "Mymensingh"

district_ts <- df %>%
  filter(District == decomp_district) %>%
  arrange(Date) %>%
  select(Date, Water_Table_m)

# Interpolate to regular interval for decomposition
# (Decomposition works better with regular time series)
full_dates <- seq(min(district_ts$Date), max(district_ts$Date), by = "week")
interpolated <- approx(district_ts$Date, district_ts$Water_Table_m, 
                       xout = full_dates, rule = 2)
interpolated_ts <- ts(interpolated$y, frequency = 52, start = c(2015, 1))

# Decompose
decomp <- stl(interpolated_ts, s.window = "periodic")

# Plot decomposition
decomp_plot <- autoplot(decomp) +
  labs(title = paste("Seasonal Decomposition -", decomp_district))

print(decomp_plot)
```


# =========================================
# 5. DISTRIBUTION AND VARIABILITY ANALYSIS
# =========================================

## 5.1 Overall distribution

```{r}
overall_dist <- ggplot(df, aes(x = Water_Table_m)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, 
                 fill = "steelblue", alpha = 0.7) +
  geom_density(color = "red", size = 1) +
  labs(title = "Overall Distribution of Water Table Depth",
       x = "Depth (m)", y = "Density")

print(overall_dist)
```


## 5.2 Boxplot by district

```{r,fig.height=15}
boxplot_district <- ggplot(df, aes(x = reorder(District, Water_Table_m, median), 
                                    y = Water_Table_m)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.size = 0.5) +
  coord_flip() +
  labs(title = "Water Table Depth Distribution by District",
       x = "District", y = "Depth (m)") +
  theme_classic()+
  theme(axis.text.y = element_text(size = 15))
  
print(boxplot_district)
```


## 5.3 Violin plot for top 20 districts

```{r}
top_20_districts <- district_counts %>%
  head(20) %>%
  pull(District)

violin_plot <- df %>%
  filter(District %in% top_20_districts) %>%
  ggplot(aes(x = reorder(District, Water_Table_m, median), 
             y = Water_Table_m, fill = District)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  coord_flip() +
  labs(title = "Distribution of Water Table Depth - Top 20 Districts",
       x = "District", y = "Depth (m)") +
  theme_classic()+
  theme(legend.position = "none")

print(violin_plot)
```


## 5.4 Monthly boxplots (seasonal distribution)

```{r}
monthly_boxplot <- ggplot(df, aes(x = Month, y = Water_Table_m, fill = Month)) +
  geom_boxplot() +
  labs(title = "Monthly Distribution of Water Table Depth",
       x = "Month", y = "Depth (m)") +
  theme(legend.position = "none")

print(monthly_boxplot)
```


# ====================
# 6. SPATIAL ANALYSIS
# ====================

## 6.1 Map of mean water table depth

```{r}
# Note: You'll need to get actual coordinates or use district centroids
# For demonstration, I'll use approximate centroids for major cities
# (You should use the actual LATITUDE and LONGITUDE from your data)

# Check if coordinates are available
if(all(c("LATITUDE", "LONGITUDE") %in% names(df))) {
  # Use actual coordinates from data
  district_coords <- df %>%
    group_by(District) %>%
    summarise(
      lat = mean(LATITUDE, na.rm = TRUE),
      lon = mean(LONGITUDE, na.rm = TRUE),
      mean_depth = mean(Water_Table_m, na.rm = TRUE),
      sd_depth = sd(Water_Table_m, na.rm = TRUE)
    )
  
  # Create map with ggplot2
  map_plot <- ggplot(district_coords, aes(x = lon, y = lat, 
                                          color = mean_depth, size = sd_depth)) +
    geom_point(alpha = 0.8) +
    scale_color_viridis_c(name = "Mean Depth (m)", direction = -1) +
    scale_size_continuous(name = "Variability (SD)") +
    labs(title = "Spatial Distribution of Mean Water Table Depth",
         x = "Longitude", y = "Latitude") +
    theme_minimal()
  
  print(map_plot)
}
```


# ========================
# 7. CORRELATION ANALYSIS
# ========================

## 7.1 Create a regular time grid for correlation analysis

```{r}
# First, interpolate each district to a common time grid
common_dates <- seq(min(df$Date), max(df$Date), by = "month")

# Function to interpolate time series for a district
interpolate_district <- function(district_name) {
  district_data <- df %>%
    filter(District == district_name) %>%
    arrange(Date) %>%
    select(Date, Water_Table_m)
  
  # Interpolate missing values
  interp_data <- approx(district_data$Date, district_data$Water_Table_m, 
                        xout = common_dates, rule = 2)
  
  return(data.frame(
    Date = interp_data$x,
    Water_Table_m = interp_data$y,
    District = district_name
  ))
}

# Interpolate for all districts (limit to those with enough data)
districts_with_data <- district_counts %>%
  filter(n_obs >= 30) %>%
  pull(District)

interpolated_data <- map_df(districts_with_data, interpolate_district)
```


## 7.2 Correlation matrix

```{r}
# Convert to wide format
wide_data <- interpolated_data %>%
  pivot_wider(id_cols = Date, names_from = District, values_from = Water_Table_m)

# Remove date column for correlation
cor_matrix <- cor(wide_data[, -1], use = "pairwise.complete.obs")
```

```{r,fig.height=12,fig.width=15}
# Correlation heatmap
corrplot_plot <- corrplot(cor_matrix, method = "color", 
                          type = "upper", order = "hclust",
                          tl.col = "black", tl.srt = 45,
                          title = "Correlation Between Districts",
                          mar = c(0, 0, 2, 0))
```


## 7.3 Find most correlated districts

```{r}
# Convert to data frame and find top correlations
cor_df <- as.data.frame(as.table(cor_matrix))
cor_df <- cor_df %>%
  filter(as.character(Var1) != as.character(Var2)) %>%
  arrange(desc(Freq))

print("Top 10 most correlated district pairs:")
head(cor_df, 10)
```


## 7.4 Dendrogram of district similarity

```{r,fig.height=10}
library(ggdendro)
dist_matrix <- as.dist(1 - cor_matrix)
hc <- hclust(dist_matrix, method = "ward.D2")

dendro_plot <- ggdendro::ggdendrogram(hc, rotate = TRUE, size = 2) +
  labs(title = "Dendrogram of District Similarity Based on Water Table Patterns",
       x = "District", y = "Distance")

print(dendro_plot)
```


# ===========================
# 8. FUNCTIONAL DATA PREVIEW
# ===========================

## 8.1 Mean curve and functional envelope

```{r}
# Plot mean ± standard deviation for all districts
mean_curve <- interpolated_data %>%
  group_by(Date) %>%
  summarise(
    mean_depth = mean(Water_Table_m, na.rm = TRUE),
    sd_depth = sd(Water_Table_m, na.rm = TRUE)
  )

functional_preview <- ggplot(mean_curve, aes(x = Date, y = mean_depth)) +
  geom_ribbon(aes(ymin = mean_depth - sd_depth, 
                  ymax = mean_depth + sd_depth), 
              alpha = 0.3, fill = "steelblue") +
  geom_line(color = "steelblue", size = 1) +
  labs(title = "Functional Data Preview: Mean ± 1 SD",
       subtitle = "Average water table pattern across all districts",
       x = "Date", y = "Water Table Depth (m)") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5)

print(functional_preview)
```


## 8.2 Smoothed curves for selected districts

```{r}
# This gives a preview of what your fPCA will capture
library(mgcv)
```

```{r,fig.width=10}
smooth_preview <- df %>%
  filter(District %in% selected_districts) %>%
  ggplot(aes(x = Date, y = Water_Table_m, color = District)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE) +
  facet_wrap(~District, scales = "free_y", ncol = 2) +
  labs(title = "Smoothed Trends - Preview for fPCA",
       x = "Date", y = "Water Table Depth (m)") +
  theme(legend.position = "none")

print(smooth_preview)
```


# ============================
# 9. ADVANCED VISUALIZATIONS
# ============================

## 9.1 Small multiples of all districts

```{r,fig.height=10,fig.width=10}
small_multiples <- df %>%
  ggplot(aes(x = Date, y = Water_Table_m)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  facet_wrap(~District, scales = "free_y", ncol = 8) +
  labs(title = "Water Table Trends - All Districts",
       x = "Date", y = "Depth (m)") +
  theme(strip.text = element_text(size = 6),
        axis.text = element_text(size = 6))

print(small_multiples)
```


## 9.2 Heatmap of water table depth by district and year

```{r,fig.height=15}
yearly_heatmap <- df %>%
  mutate(Year = year(Date)) %>%
  group_by(District, Year) %>%
  summarise(mean_depth = mean(Water_Table_m, na.rm = TRUE)) %>%
  filter(Year >= 2015 & Year <= 2024) %>%
  ggplot(aes(x = Year, y = reorder(District, mean_depth), fill = mean_depth)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Depth (m)", direction = -1) +
  labs(title = "Yearly Average Water Table Depth by District",
       x = "Year", y = "District") +
  theme(axis.text.y = element_text(size = 12))

print(yearly_heatmap)
```

