---
title: "Clustering"
author: "**Istiak Ahmed***, SO, ASICT Division, BARI*"
output: html_notebook
---

```{r}
rm(list = ls())
```

```{r}
library(fda)
```

```{r}
wt <- read.csv("district_monthly_avg_wt.csv")
```

```{r}
wt_mat <- as.matrix(wt[ , -1])
```

```{r}
wt_basis <- create.fourier.basis(c(0,12),12)
harmLfd = vec2Lfd(c(0,(2*pi/12)^2,0), c(0, 12))
```

```{r}
wt_fdPar <- fdPar(wt_basis,harmLfd,1e2)
wt_fd <- smooth.basis(1:12, wt_mat, wt_fdPar)
```

## Functional curves

```{r,fig.width=10}
plot(wt_fd$fd,
     xlab="Time (Months)",
     ylab="Water Table Depth",
     lwd=2)
```

## Variance-Covariance Surface

```{r}
wt_var <- var.fd(wt_fd$fd)
wt_vals <- eval.bifd(1:12, 1:12, wt_var)
```

```{r}
contour(1:12, 1:12, wt_vals,
        xlab="Month", ylab="Month")
```

## Correlation Surface

```{r}
wt_cor <- cor.fd(1:12, wt_fd$fd)
```

```{r}
contour(1:12, 1:12, wt_cor,
           xlab="Month", ylab="Month")
```

```{r}
wt_pca <- pca.fd(wt_fd$fd, nharm=2)
```

```{r}
wt_pca$varprop
```

## Scree-plot

```{r}
plot(wt_pca$values,
     type="b",
     pch=19,
     xlab="Component",
     ylab="Eigenvalue")
```

```{r}
#temppca$values are the eigenvalues
par(mfrow=c(1,1),mar = c(8, 8, 4, 2))
plot(wt_pca$values[1:8],xlab='component',ylab='variance',col="red",
     cex.lab=1.5,cex.axis=1.5,cex=2)

# plot the cumulative percentage explained total variations
# It shows that the top 3 FPCs explains more than 99% of total variations
par(mfrow=c(1,1),mar = c(8, 8, 4, 2))
plot(cumsum(wt_pca$values[1:10])/sum(wt_pca$values),xlab='Number of Components',
     ylab='cumulative variance explained')
abline(h=0.995)
```

```{r}
harmvals <- eval.fd(1:12, wt_pca$harmonics)
```

## PC's Plot

```{r,fig.width=10}
matplot(1:12, harmvals,
        type="l",
        lwd=3,
        xlab="Month",
        ylab="FPCs")

legend("topright",
       legend=paste0("PC",1:2),
       col=1:2,
       lty=1,
       lwd=3)
```

## plot the first FPC scores vs. the second FPC scores

```{r,fig.height=10,fig.width=15}
plot(wt_pca$scores[,1],
     wt_pca$scores[,2],
     pch=19,
     xlab="PC1",
     ylab="PC2")

text(wt_pca$scores[,1],
     wt_pca$scores[,2],
     labels=colnames(wt_mat),
     pos=3)
```

```{r}
pc_scores <- wt_pca$scores
district_names <- colnames(wt_mat)
```

```{r}
vulnerability <- data.frame(
  District = district_names,
  PC1 = pc_scores[,1],
  PC2 = pc_scores[,2]
)
```

```{r}
vulnerability <- vulnerability[order(vulnerability$PC1), ]
```

```{r}
vulnerability
```

## Clustering the districts based on vulnerability

```{r}
scores <- wt_pca$scores[,1:2]
```

```{r}
wss <- numeric(10)
```

```{r}
for(i in 1:10){
  wss[i] <- kmeans(scores, centers=i)$tot.withinss
}
```

```{r}
plot(1:10, wss, type="b",
     xlab="Number of Clusters",
     ylab="Within-cluster sum of squares")
```

```{r}
set.seed(3287)  # for reproducibility
km <- kmeans(scores, centers = 4, nstart = 25)
clusters <- km$cluster
```

```{r}
cluster_result <- data.frame(
  District = colnames(wt_mat),
  PC1 = scores[,1],
  PC2 = scores[,2],
  Cluster = clusters
)
```


```{r}
library(car)
```

```{r,fig.height=9}
plot(scores[,1], scores[,2],
     col = clusters,
     pch = 19,
     xlab = "PC1",
     ylab = "PC2")
text(scores[,1], scores[,2],
     labels = colnames(wt_mat),
     pos = 3)

for(k in sort(unique(clusters))){
  pts <- scores[clusters == k, , drop = FALSE]
  if(nrow(pts) >= 3){  # need >=3 points for a covariance ellipse
    car::dataEllipse(pts[,1], pts[,2],
                     levels = 0.68,        # ~3 SD ellipsoid
                     add = TRUE,
                     col = k,
                     center.pch = FALSE,
                     plot.points = FALSE,
                     fill = TRUE, fill.alpha = 0.1)
  }
}
```

```{r,fig.height=10,fig.width=15}
plot(scores[,1], scores[,2],
     col = clusters,
     pch = 19,
     xlab = "PC1",
     ylab = "PC2")

text(scores[,1], scores[,2],
     labels = colnames(wt_mat),
     pos = 3)
```

```{r}
# water level statistics for each district
water_stats <- data.frame(
  District = colnames(wt_mat),
  Mean = apply(wt_mat, 2, mean, na.rm = TRUE),
  Max = apply(wt_mat, 2, max, na.rm = TRUE),
  Min = apply(wt_mat, 2, min, na.rm = TRUE)
)

# merge with cluster results
cluster_table <- merge(cluster_result, water_stats, by="District")

# sort by cluster
cluster_table <- cluster_table[order(cluster_table$Cluster), ]

cluster_table
```

```{r}
cluster_desc = aggregate(cbind(Mean, Max, Min) ~ Cluster, data = cluster_table, mean)
print(cluster_desc)
```

```{r}
split(cluster_table$District, cluster_table$Cluster)
```

| Cluster | Suggested Name | Interpretation |
|----------------|----------------------------|----------------------------|
| Cluster 1 | **Stable Groundwater Regime** | Districts with relatively stable groundwater levels and low seasonal variability. |
| Cluster 2 | **Moderately Fluctuating Groundwater Regime** | Districts with moderate seasonal fluctuations but generally balanced recharge and withdrawal. |
| Cluster 3 | **Seasonally Stressed Groundwater Regime** | Districts showing strong seasonal depletion, indicating high dependence on monsoon recharge. |
| Cluster 4 | **Highly Vulnerable Groundwater Regime** | Districts with deeper groundwater levels and strong fluctuations, indicating potential over-extraction or limited recharge. |


# ==============================================
# Formal Cluster Validity — Silhouette + NbClust
# ==============================================

```{r}
library(cluster)
library(factoextra)
library(NbClust)
```


## ---------------------------------------
## 1A. Silhouette plot for your chosen k=4
## ---------------------------------------

```{r}
sil <- silhouette(clusters, dist(scores))
```

```{r}
# Summary: average silhouette width per cluster
sil_summary <- summary(sil)
cat("Average silhouette width (k=4):", round(sil_summary$avg.width, 4), "\n")
cat("Silhouette width by cluster:\n")
print(round(sil_summary$clus.avg.widths, 4))
```

```{r}
# Publication-quality silhouette plot
fviz_silhouette(sil,
                palette     = c("#1B6CA8", "#16A085", "#E67E22", "#8E44AD"),
                ggtheme     = theme_classic(),
                print.summary = FALSE) +
  labs(title    = "Silhouette Plot for k = 4 Clusters",
       subtitle = paste0("Average silhouette width = ",
                         round(sil_summary$avg.width, 3)),
       x = "Districts (ordered by cluster)",
       y = "Silhouette width") +
  theme(axis.text.x  = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title   = element_text(size = 13, face = "bold"))
```

## -----------------------------------------------
## 1B. Compare silhouette width across k = 2 to 8
## -----------------------------------------------

```{r}
avg_sil <- numeric(8)
avg_sil[1] <- NA  # k=1 not defined

for (k in 2:8) {
  hc_k   <- hclust(dist(scores), method = "ward.D2")
  cl_k   <- cutree(hc_k, k = k)
  sil_k  <- silhouette(cl_k, dist(scores))
  avg_sil[k] <- mean(sil_k[, 3])
}

sil_df <- data.frame(k = 2:8, avg_sil = avg_sil[2:8])
```

```{r}
ggplot(sil_df, aes(x = k, y = avg_sil)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(size = 3, color = "steelblue") +
  geom_point(data = sil_df[sil_df$k == 4, ],
             aes(x = k, y = avg_sil),
             size = 5, shape = 21,
             fill = "#E67E22", color = "#E67E22") +
  annotate("text", x = 4.2, y = sil_df$avg_sil[sil_df$k == 4],
           label = paste0("k=4\n(", round(sil_df$avg_sil[sil_df$k == 4], 3), ")"),
           hjust = 0, size = 3.5, color = "#E67E22") +
  scale_x_continuous(breaks = 2:8) +
  labs(title    = "Average Silhouette Width by Number of Clusters",
       subtitle = "Higher values indicate better-defined clusters",
       x = "Number of clusters (k)",
       y = "Average silhouette width") +
  theme_classic() +
  theme(plot.title = element_text(size = 13, face = "bold"))
```

## ----------------------------------------------
## 1C. NbClust — majority vote across 26 indices
## ----------------------------------------------

```{r}
nb <- NbClust(data    = scores,
              diss    = NULL,
              distance = "euclidean",
              min.nc  = 2,
              max.nc  = 8,
              method  = "ward.D2",
              index   = "all")
```


```{r}
cat("\nNbClust best number of clusters:", nb$Best.nc[1, which.max(nb$Best.nc[2,])], "\n")
```


# ============================================================
# JUSTIFYING k=4 OVER k=3 — Complete Argumentation Code
# Provides statistical + substantive evidence for 4 clusters
# ============================================================

```{r}
library(cluster)
library(factoextra)
library(ggplot2)
library(patchwork)
library(dplyr)
```


## ----------------------------------------------------------
## ARGUMENT 1: NbClust vote breakdown — k=4 is competitive
## The majority rule is misleading when votes are split.
## Visualise the actual vote distribution honestly.
## ----------------------------------------------------------

```{r}
# Recreate vote counts from your NbClust output
vote_df <- data.frame(
  k     = c(2, 3, 4, 5, 6, 7, 8),
  votes = c(6, 6, 4, 1, 4, 1, 1)
)

p_votes <- ggplot(vote_df, aes(x = factor(k), y = votes,
                                fill = factor(k == 4))) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = votes), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#E67E22"),
                    guide = "none") +
  annotate("text", x = "4", y = 4.6,
           label = "chosen", color = "#E67E22",
           size = 3.5, fontface = "italic") +
  labs(
    title    = "NbClust: Votes per Number of Clusters",
    subtitle = paste0(
      "k=2 and k=3 tied at 6 votes each; k=4 received 4 votes.\n",
      "No single k dominates \u2014 majority rule alone is insufficient."
    ),
    x = "Number of clusters (k)",
    y = "Number of indices recommending k"
  ) +
  theme_classic() +
  theme(plot.title    = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 9, color = "grey30"))
```

```{r}
print(p_votes)
```

## ----------------------------------------------------------
## ARGUMENT 2: Silhouette comparison k=3 vs k=4
## Show that the DROP in silhouette from k=3 to k=4 is small.
## ----------------------------------------------------------

```{r}
sil_vals <- sapply(2:6, function(k) {
  cl  <- cutree(hclust(dist(scores), method = "ward.D2"), k = k)
  mean(silhouette(cl, dist(scores))[, 3])
})

sil_comp <- data.frame(k = 2:6, avg_sil = sil_vals)
```

```{r}
p_sil <- ggplot(sil_comp, aes(x = factor(k), y = avg_sil,
                               fill = factor(k %in% c(3, 4)))) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = round(avg_sil, 3)), vjust = -0.4, size = 4) +
  scale_fill_manual(
    values = c("FALSE" = "grey80", "TRUE" = "#1B6CA8"),
    guide  = "none"
  ) +
  labs(
    title    = "Average Silhouette Width: k=3 vs k=4",
    subtitle = "Small decrease from k=3 to k=4 is offset by substantive gain",
    x = "Number of clusters (k)",
    y = "Average silhouette width"
  ) +
  theme_classic() +
  theme(plot.title    = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 9, color = "grey30"))

```

```{r}
print(p_sil)
```

```{r}
# Print the actual drop
sil_k3 <- sil_comp$avg_sil[sil_comp$k == 3]
sil_k4 <- sil_comp$avg_sil[sil_comp$k == 4]
cat(sprintf(
  "Silhouette: k=3 = %.4f, k=4 = %.4f, drop = %.4f (%.1f%%)\n",
  sil_k3, sil_k4, sil_k3 - sil_k4,
  (sil_k3 - sil_k4) / sil_k3 * 100
))
```

## ----------------------------------------------------------
## ARGUMENT 3: What does k=3 actually merge?
## Show that k=3 collapses Clusters 3 and 4 into one group,
## hiding the hydrogeologically distinct deep-water districts.
## ----------------------------------------------------------

```{r}
clusters_k3 <- cutree(hclust(dist(scores), method = "ward.D2"), k = 3)
clusters_k4 <- clusters   # your existing k=4 result

comp_df <- data.frame(
  District  = colnames(wt_mat),
  Cluster_3 = clusters_k3,
  Cluster_4 = clusters_k4,
  Mean_WT   = apply(wt_mat, 2, mean, na.rm = TRUE),
  Max_WT    = apply(wt_mat, 2, max,  na.rm = TRUE)
)
```

```{r}
# Which k=4 clusters get merged by k=3?
cat("\n=== How k=3 merges the k=4 clusters ===\n")
print(table(k3 = comp_df$Cluster_3, k4 = comp_df$Cluster_4))

# Identify the merged cluster (the one that combines k=4 Clusters 3 & 4)
merged <- comp_df %>%
  filter(Cluster_4 %in% c(3, 4)) %>%
  arrange(Cluster_4, desc(Mean_WT))

cat("\n=== Districts merged by k=3 (k=4 Clusters 3 & 4 combined) ===\n")
cat("These districts have very different mean water table depths:\n\n")
print(merged[, c("District", "Cluster_3", "Cluster_4", "Mean_WT", "Max_WT")])

# Summary of what gets merged
cat("\n--- Mean water table depth: k=4 Cluster 3 vs Cluster 4 ---\n")
agg <- aggregate(Mean_WT ~ Cluster_4,
                 data = comp_df[comp_df$Cluster_4 %in% c(3, 4), ],
                 FUN  = function(x) round(mean(x), 3))
print(agg)
cat("=> k=3 would average these two hydrogeologically distinct groups.\n")
```

## ----------------------------------------------------------
## ARGUMENT 4: Visual — PC score biplot showing k=3 vs k=4
## Side by side to show what gets lost with k=3
## ----------------------------------------------------------

```{r}
score_df <- data.frame(
  District  = colnames(wt_mat),
  PC1       = scores[, 1],
  PC2       = scores[, 2],
  Cluster_3 = factor(clusters_k3),
  Cluster_4 = factor(clusters_k4)
)
```

```{r}
# k=3 biplot
p_k3 <- ggplot(score_df, aes(x = PC1, y = PC2, color = Cluster_3)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = District),
                            size = 2.5, max.overlaps = 20,
                            segment.color = "grey70") +
  scale_color_manual(values = c("#1B6CA8", "#16A085", "#E67E22"),
                     name = "Cluster") +
  labs(title = "k = 3",
       subtitle = paste0("Avg silhouette = ", round(sil_k3, 3)),
       x = "PC1 (94.1%)", y = "PC2 (4.5%)") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom")

# k=4 biplot
p_k4 <- ggplot(score_df, aes(x = PC1, y = PC2, color = Cluster_4)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = District),
                            size = 2.5, max.overlaps = 20,
                            segment.color = "grey70") +
  scale_color_manual(values = c("#1B6CA8", "#16A085", "#E67E22", "#8E44AD"),
                     name = "Cluster") +
  labs(title = "k = 4  (chosen)",
       subtitle = paste0("Avg silhouette = ", round(sil_k4, 3)),
       x = "PC1 (94.1%)", y = "PC2 (4.5%)") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom")

(p_k3 | p_k4) +
  plot_annotation(
    title   = "Cluster Solution Comparison: k=3 vs k=4",
    caption = paste0(
      "k=4 separates the highly vulnerable deep-water districts (Mymensingh, Naogaon, ",
      "Nawabganj, Moulvibazar)\nfrom the seasonally stressed group \u2014 ",
      "a distinction with clear hydrogeological significance."
    ),
    theme = theme(
      plot.title   = element_text(size = 13, face = "bold"),
      plot.caption = element_text(size = 8.5, color = "grey30")
    )
  )
```

## ----------------------------------------------------------
## ARGUMENT 5: Water table depth profiles by cluster
## Show that Cluster 4 has a distinctly different profile
## that justifies its separation from Cluster 3
## ----------------------------------------------------------

```{r}
# Long format for plotting monthly profiles by cluster
wt_long <- as.data.frame(t(wt_mat)) %>%
  tibble::rownames_to_column("District") %>%
  tidyr::pivot_longer(-District, names_to = "Month_num",
                      values_to = "WaterTable") %>%
  mutate(Month_num = as.integer(gsub("V", "", Month_num))) %>%
  left_join(comp_df[, c("District", "Cluster_4")], by = "District") %>%
  filter(!is.na(WaterTable))
```

```{r}
# Mean profile per cluster
cluster_means <- wt_long %>%
  group_by(Cluster_4, Month_num) %>%
  summarise(mean_wt = mean(WaterTable, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Cluster_label = recode(as.character(Cluster_4),
      "1" = "C1: Stable",
      "2" = "C2: Moderately Fluctuating",
      "3" = "C3: Seasonally Stressed",
      "4" = "C4: Highly Vulnerable"
    )
  )
```

```{r}
ggplot(cluster_means, aes(x = Month_num, y = mean_wt,
                           color = Cluster_label,
                           linetype = Cluster_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_manual(values = c(
    "C1: Stable"                  = "#1B6CA8",
    "C2: Moderately Fluctuating"  = "#16A085",
    "C3: Seasonally Stressed"     = "#E67E22",
    "C4: Highly Vulnerable"       = "#8E44AD"
  ), name = NULL) +
  scale_linetype_manual(values = c(
    "C1: Stable"                  = "solid",
    "C2: Moderately Fluctuating"  = "dashed",
    "C3: Seasonally Stressed"     = "dotdash",
    "C4: Highly Vulnerable"       = "dotted"
  ), name = NULL) +
  labs(
    title    = "Mean Monthly Water Table Depth by Cluster (k=4)",
    subtitle = paste0(
      "Cluster 4 is clearly separated from Cluster 3 in both depth and amplitude.\n",
      "Merging them (as k=3 would) obscures a critical vulnerability distinction."
    ),
    x = "Month", y = "Mean Water Table Depth (m)"
  ) +
  theme_classic() +
  theme(
    plot.title      = element_text(size = 12, face = "bold"),
    plot.subtitle   = element_text(size = 9,  color = "grey30"),
    legend.position = "bottom",
    axis.text.x     = element_text(angle = 45, hjust = 1)
  )
```

## ----------------------------------------------------------
## PRINT: Ready-to-use justification paragraph for the paper
## ----------------------------------------------------------

```{r}
cat("\n")
cat("=====================================================\n")
cat("SUGGESTED JUSTIFICATION TEXT FOR METHODS/RESULTS:\n")
cat("=====================================================\n")
cat(sprintf(
"The NbClust procedure applied to the FPC scores yielded no clear
consensus: k=2 and k=3 each received 6 votes from 23 applicable indices,
while k=4 received 4 votes (k=6 also received 4). Given this ambiguity,
the final cluster number was determined by combining three criteria.
First, the within-cluster sum of squares (elbow plot) showed a discernible
reduction in marginal gain beyond k=4. Second, the average silhouette
width for k=4 (%.3f) was only marginally lower than for k=3 (%.3f),
a difference of %.3f, indicating that the additional cluster does not
substantially reduce compactness. Third, and most critically, the k=3
solution collapsed two hydrogeologically distinct groups — the seasonally
stressed districts (mean depth: %.1fm) and the highly vulnerable deep-water
districts (mean depth: %.1fm) — into a single cluster, obscuring a
meaningful difference of %.1fm in mean water table depth. This deep-water
group (Mymensingh, Naogaon, Nawabganj, Moulvibazar) corresponds to the
Barind Tract and Haor Basin regions, which are known for their
exceptionally deep groundwater and represent a qualitatively different
vulnerability regime. Therefore, k=4 was retained as the optimal solution
on grounds of both statistical adequacy and hydrogeological interpretability.\n",
  sil_k4, sil_k3, sil_k3 - sil_k4,
  mean(comp_df$Mean_WT[comp_df$Cluster_4 == 3]),
  mean(comp_df$Mean_WT[comp_df$Cluster_4 == 4]),
  mean(comp_df$Mean_WT[comp_df$Cluster_4 == 4]) -
    mean(comp_df$Mean_WT[comp_df$Cluster_4 == 3])
))
```

## ----------------------------------------------------------
## 2A. Evaluate mean curve and harmonics on a fine grid
## ----------------------------------------------------------

```{r}
t_fine   <- seq(1, 12, length.out = 200)   # fine monthly grid

# Mean function
mean_vals <- eval.fd(t_fine, mean.fd(wt_fd$fd))

# Harmonic functions (PC1 and PC2)
harm_vals <- eval.fd(t_fine, wt_pca$harmonics)

# Scaling factor: typically 2 * sqrt(eigenvalue) gives visible perturbation
scale1 <- 2 * sqrt(wt_pca$values[1])
scale2 <- 2 * sqrt(wt_pca$values[2])

# Plus and minus perturbations
pc1_plus  <- mean_vals + scale1 * harm_vals[, 1]
pc1_minus <- mean_vals - scale1 * harm_vals[, 1]
pc2_plus  <- mean_vals + scale2 * harm_vals[, 2]
pc2_minus <- mean_vals - scale2 * harm_vals[, 2]
```

## ----------------------------------------------------------
## 2B. Build data frames for ggplot
## ----------------------------------------------------------

```{r}
month_labels <- month.abb  # Jan–Dec

df_pc1 <- data.frame(
  t      = rep(t_fine, 3),
  value  = c(mean_vals, pc1_plus, pc1_minus),
  curve  = rep(c("Mean", "+PC1", "\u2212PC1"), each = length(t_fine))
)

df_pc2 <- data.frame(
  t      = rep(t_fine, 3),
  value  = c(mean_vals, pc2_plus, pc2_minus),
  curve  = rep(c("Mean", "+PC2", "\u2212PC2"), each = length(t_fine))
)
```

## ----------------------------------------------------------
## 2C. Plot PC1 effect
## ----------------------------------------------------------

```{r}
p1 <- ggplot(df_pc1, aes(x = t, y = value,
                          color = curve, linetype = curve)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Mean"  = "black",
                                "+PC1"  = "#1B6CA8",
                                "\u2212PC1" = "#C0392B"),
                     name = NULL) +
  scale_linetype_manual(values = c("Mean"  = "solid",
                                   "+PC1"  = "dashed",
                                   "\u2212PC1" = "dotted"),
                        name = NULL) +
  scale_x_continuous(breaks = 1:12, labels = month_labels) +
  labs(
    title    = paste0("PC1 Effect  (",
                      round(wt_pca$varprop[1] * 100, 1), "% variance explained)"),
    subtitle = "Mean curve \u00b1 2\u221a\u03bb\u2081 \u00d7 PC1 harmonic",
    x = "Month", y = "Water Table Depth (m)"
  ) +
  theme_classic() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    axis.text.x  = element_text(angle = 45, hjust = 1)
  )

print(p1)
```

## ----------------------------------------------------------
## 2D. Plot PC2 effect
## ----------------------------------------------------------

```{r}
p2 <- ggplot(df_pc2, aes(x = t, y = value,
                          color = curve, linetype = curve)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Mean"  = "black",
                                "+PC2"  = "#16A085",
                                "\u2212PC2" = "#8E44AD"),
                     name = NULL) +
  scale_linetype_manual(values = c("Mean"  = "solid",
                                   "+PC2"  = "dashed",
                                   "\u2212PC2" = "dotted"),
                        name = NULL) +
  scale_x_continuous(breaks = 1:12, labels = month_labels) +
  labs(
    title    = paste0("PC2 Effect  (",
                      round(wt_pca$varprop[2] * 100, 1), "% variance explained)"),
    subtitle = "Mean curve \u00b1 2\u221a\u03bb\u2082 \u00d7 PC2 harmonic",
    x = "Month", y = "Water Table Depth (m)"
  ) +
  theme_classic() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    axis.text.x  = element_text(angle = 45, hjust = 1)
  )

print(p2)
```

## ----------------------------------------------------------
## 2E. Combined panel (publication layout)
## ----------------------------------------------------------

```{r,fig.width=9}
library(patchwork)
(p1 | p2) +
  plot_annotation(
    title   = "Functional Principal Components: Temporal Effects on Water Table",
    caption = paste0(
      "PC1 explains ", round(wt_pca$varprop[1] * 100, 1), "% of variance; ",
      "PC2 explains ", round(wt_pca$varprop[2] * 100, 1), "%. ",
      "Dashed/dotted lines show effect of adding/subtracting 2\u221a\u03bb \u00d7 PC harmonic from the mean."
    ),
    theme = theme(
      plot.title   = element_text(size = 14, face = "bold"),
      plot.caption = element_text(size = 9, color = "gray40")
    )
  )
```

## ----------------------------------------------------------
## 2F. Interpretation helper — print what the PCs mean
## ----------------------------------------------------------

```{r}
# Evaluate harmonics at key months to help narrate
harm_at_months <- eval.fd(1:12, wt_pca$harmonics)
cat("\n--- PC1 harmonic values by month (interpret sign & magnitude) ---\n")
pc1_by_month <- data.frame(
  Month = month.abb,
  PC1_harmonic = round(harm_at_months[, 1], 4),
  PC2_harmonic = round(harm_at_months[, 2], 4)
)
print(pc1_by_month)

cat("\nInterpretation guide:\n")
cat("  High PC1 score => district's curve is shifted UP relative to mean\n")
cat("  (water table deeper = more vulnerable in dry season)\n")
cat("  Positive PC1 harmonic months show where this separation is largest.\n")
```

