---
title: "R Notebook"
author: "**Istiak Ahmed***, SO, ASICT Division, BARI*"
output: html_notebook
---

## ============================================================
## Strategy: train on 2015–2021 (7 years), test on 2022–2024 (3 years)
### Apply to all 4 districts consistently
## ============================================================

```{r}
library(ftsa)
```

## ----------------------------------------------------------
## Helper: compute accuracy metrics for one district
## ----------------------------------------------------------

```{r}
backtest_district <- function(csv_file, district_name,
                               train_years = 7, h = 3,
                               order = 3) {

  df    <- read.csv(csv_file)
  y_mat <- as.matrix(df)
  colnames(y_mat) <- as.character(2015:(2015 + ncol(y_mat) - 1))

  n_total <- ncol(y_mat)          # should be 10
  n_train <- train_years          # 8
  n_test  <- n_total - n_train    # 2

  # Split
  y_train <- y_mat[, 1:n_train]
  y_test  <- y_mat[, (n_train + 1):n_total]

  x <- 1:nrow(y_mat)  # months 1–12

  # Fit on training set
  fts_train <- fts(x      = x,
                   y      = y_train,
                   xname  = "Month",
                   yname  = "Water Table Depth")

  fit <- ftsm(fts_train, order = order, method = "rapca")

  # Forecast h years ahead
  fc  <- forecast(fit, h = n_test, method = "arfima")

  # Predicted values: fc$mean$y is a 12 x h matrix
  y_pred <- fc$mean$y   # rows = months, cols = forecast years

  # Actual test values
  y_actual <- y_test    # 12 x 2 matrix

  # Flatten for metrics
  errors   <- as.vector(y_actual) - as.vector(y_pred)
  rmse     <- sqrt(mean(errors^2, na.rm = TRUE))
  mae      <- mean(abs(errors),   na.rm = TRUE)
  mape     <- mean(abs(errors / as.vector(y_actual)) * 100, na.rm = TRUE)

  # Monthly RMSE (useful to see which months are hardest to predict)
  monthly_rmse <- apply((y_actual - y_pred)^2, 1, function(x) sqrt(mean(x, na.rm = TRUE)))

  list(
    district     = district_name,
    rmse         = round(rmse, 4),
    mae          = round(mae, 4),
    mape         = round(mape, 2),
    monthly_rmse = round(monthly_rmse, 4),
    y_actual     = y_actual,
    y_pred       = y_pred,
    y_train      = y_train,
    x            = x,
    varprop      = fit$varprop
  )
}
```

## ----------------------------------------------------------
## Run for all 4 districts
### Adjust file paths to match your working directory
## ----------------------------------------------------------

```{r}
results <- list(
  backtest_district("Borguna.csv",    "Barguna"),
  backtest_district("Joypurhat.csv",  "Joypurhat"),
  backtest_district("Mymensingh.csv", "Mymensingh"),
  backtest_district("Netrokona.csv",  "Netrokona",  order = 5)
)
```

#### Note: Netrokona uses order=5 due to its lower PC1 variance explained (37.8%)

## ----------------------------------------------------------
## Accuracy table (for the paper)
## ----------------------------------------------------------

```{r}
acc_table <- do.call(rbind, lapply(results, function(r) {
  data.frame(District = r$district,
             RMSE     = r$rmse,
             MAE      = r$mae,
             MAPE     = paste0(r$mape, "%")
             )
}))
```

```{r}
print("=== Forecast Accuracy (2-year holdout: 2023–2024) ===")
print(acc_table, row.names = FALSE)
```

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
```

```{r}
acc_long <- acc_table %>%
  select(District, RMSE, MAE) %>%
  pivot_longer(cols = c(RMSE, MAE), names_to = "Metric", values_to = "Value")
```


```{r}
ggplot(acc_long, aes(x = District, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("RMSE" = "#1B6CA8", "MAE" = "#16A085")) +
  labs(title    = "Forecast Accuracy by District (2-Year Holdout)",
       subtitle = "Training: 2015–2022 | Test: 2023–2024",
       x = NULL, y = "Error (metres)", fill = NULL) +
  theme_classic() +
  theme(plot.title    = element_text(size = 13, face = "bold"),
        legend.position = "top")
```



## ----------------------------------------------------------
## Visual: Actual vs Predicted for each district
## ----------------------------------------------------------

```{r,fig.height=10,fig.width=10}
month_labels <- month.abb

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (r in results) {
  # Training years on x-axis, then test years
  n_train <- ncol(r$y_train)

  # Plot all observed years (grey)
  matplot(r$x, r$y_train,
          type = "l", col = "grey70", lty = 1, lwd = 1,
          xlab = "Month", ylab = "Water Table Depth (m)",
          main = r$district,
          xaxt = "n", ylim = range(c(r$y_train, r$y_actual, r$y_pred), na.rm = TRUE))
  axis(1, at = 1:12, labels = month_labels, las = 2, cex.axis = 0.7)

  # Actual test years (solid dark)
  matlines(r$x, r$y_actual, col = c("black", "black"), lty = c(1, 2), lwd = 2)

  # Predicted test years (coloured)
  matlines(r$x, r$y_pred, col = c("#C0392B", "#C0392B"), lty = c(1, 2), lwd = 2)

  legend("topright",
         legend = c("Training", "Actual 2023", "Actual 2024",
                    "Forecast 2023", "Forecast 2024"),
         col    = c("grey70", "black", "black", "#C0392B", "#C0392B"),
         lty    = c(1, 1, 2, 1, 2),
         lwd    = c(1, 2, 2, 2, 2),
         cex    = 0.6, bty = "n")

  mtext(paste0("RMSE=", r$rmse, "  MAE=", r$mae), side = 3,
        line = -1.2, cex = 0.7, col = "grey30")
}
```


## ----------------------------------------------------------
## Monthly RMSE heatmap (shows which months are hardest)
## ----------------------------------------------------------

```{r}
monthly_df <- do.call(rbind, lapply(results, function(r) {
  data.frame(
    District = r$district,
    Month    = month.abb,
    RMSE     = r$monthly_rmse
  )
}))
monthly_df$Month <- factor(monthly_df$Month, levels = month.abb)
```


```{r}
ggplot(monthly_df, aes(x = Month, y = District, fill = RMSE)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(RMSE, 2)), size = 3) +
  scale_fill_gradient(low = "#EAF3DE", high = "#A32D2D",
                      name = "RMSE (m)") +
  labs(title    = "Monthly Forecast RMSE by District",
       subtitle = "Higher values indicate months with greater forecast error",
       x = NULL, y = NULL) +
  theme_classic() +
  theme(axis.text.x   = element_text(angle = 45, hjust = 1),
        plot.title     = element_text(size = 13, face = "bold"))
```


## ----------------------------------------------------------
## District configuration
### Use order=5 for Netrokona (low PC1 variance), 3 for others
## ----------------------------------------------------------

```{r}
district_configs <- list(
  list(file = "Borguna.csv",    name = "Barguna",    order = 3),
  list(file = "Joypurhat.csv",  name = "Joypurhat",  order = 3),
  list(file = "Mymensingh.csv", name = "Mymensingh", order = 3),
  list(file = "Netrokona.csv",  name = "Netrokona",  order = 5)
)
```

```{r}
train_years    <- 2015:2024
forecast_years <- 2025:2030
month_labels   <- month.abb
```
 
## ----------------------------------------------------------
## Fit and forecast for each district
## ----------------------------------------------------------

```{r}
full_forecasts <- lapply(district_configs, function(cfg) {
 
  df    <- read.csv(cfg$file)
  y_mat <- as.matrix(df)
  colnames(y_mat) <- as.character(train_years)
  x <- 1:nrow(y_mat)
 
  fts_obj <- fts(x      = x,
                 y      = y_mat,
                 xname  = "Month",
                 yname  = "Water Table Depth")
 
  fit <- ftsm(fts_obj, order = cfg$order, method = "rapca")
  fc  <- forecast(fit, h = 6, method = "arfima")
 
  # fc$mean$y is a 12 x 6 matrix (rows=months, cols=forecast years)
  fc_mat <- fc$mean$y
  colnames(fc_mat) <- as.character(forecast_years)
 
  list(
    name      = cfg$name,
    order     = cfg$order,
    varprop   = fit$varprop,
    y_train   = y_mat,
    fc_mat    = fc_mat
  )
})
```

## ----------------------------------------------------------
## Build tidy long-format data for plotting
## ----------------------------------------------------------

```{r}
build_plot_data <- function(res) {
 
  # Historical
  hist_df       <- as.data.frame(res$y_train)
  hist_df$Month <- 1:12
  hist_long     <- pivot_longer(hist_df, -Month,
                                names_to  = "Year",
                                values_to = "Value") %>%
    mutate(Year   = as.integer(Year),
           Series = "Historical")
 
  # Forecast
  fc_df       <- as.data.frame(res$fc_mat)
  fc_df$Month <- 1:12
  fc_long     <- pivot_longer(fc_df, -Month,
                              names_to  = "Year",
                              values_to = "Value") %>%
    mutate(Year   = as.integer(Year),
           Series = "Forecast")
 
  # Stitch last observed year to first forecast year for visual continuity
  bridge <- hist_long %>%
    filter(Year == max(train_years)) %>%
    mutate(Series = "Forecast")
 
  bind_rows(hist_long, bridge, fc_long)
}
```

```{r}
all_plot_data <- lapply(full_forecasts, build_plot_data)
```

## ----------------------------------------------------------
## PLOT 1: Faceted by month (12-panel grid per district)
### Shows how each calendar month evolves over 2015–2030
## ----------------------------------------------------------

```{r}
plot_faceted <- function(res, dat) {
 
  ggplot(dat, aes(x = Year, y = Value, color = Series, linetype = Series)) +
    geom_line(linewidth = 0.8) +
    geom_vline(xintercept = 2024.5,
               linetype = "dotted", color = "grey40", linewidth = 0.5) +
    facet_wrap(~ Month, nrow = 3, ncol = 4,
               labeller = labeller(Month = setNames(month_labels, 1:12))) +
    scale_color_manual(
      values = c("Historical" = "grey40", "Forecast" = "#C0392B"),
      name   = NULL
    ) +
    scale_linetype_manual(
      values = c("Historical" = "solid", "Forecast" = "dashed"),
      name   = NULL
    ) +
    scale_x_continuous(
      breaks = c(2015, 2018, 2021, 2024, 2027, 2030),
      labels = c("2015","2018","2021","2024","2027","2030")
    ) +
    labs(
      title    = paste(res$name,
                       "\u2014 Monthly Water Table Forecast (2025\u20132030)"),
      subtitle = paste0(
        "Training: 2015\u20132024 (full 10 years)  |  ",
        "Model: FTSM (RAPCA, order = ", res$order, ", ARFIMA)  |  ",
        "PC1 variance explained: ",
        round(res$varprop[1] * 100, 1), "%"
      ),
      x = "Year",
      y = "Water Table Depth (m)"
    ) +
    theme_classic() +
    theme(
      plot.title      = element_text(size = 12, face = "bold"),
      plot.subtitle   = element_text(size = 8,  color = "grey30"),
      strip.text      = element_text(size = 9,  face = "bold"),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
      axis.text.y     = element_text(size = 7),
      legend.position = "bottom"
    )
}
```

## ----------------------------------------------------------
## PLOT 2: Selected months overlay (Jan, Apr, Jul, Oct)
### Shows seasonal spread across years on a single panel
## ----------------------------------------------------------

```{r}
plot_selected <- function(res, dat,
                           selected = c(1, 4, 7, 10)) {
 
  dat_sel <- dat %>%
    filter(Month %in% selected) %>%
    mutate(Month_label = factor(month_labels[Month],
                                levels = month_labels[selected]))
 
  colors <- c("#1B6CA8", "#16A085", "#E67E22", "#8E44AD")
  names(colors) <- month_labels[selected]
 
  ggplot(dat_sel,
         aes(x = Year, y = Value,
             color = Month_label, linetype = Series)) +
    geom_line(linewidth = 1) +
    geom_vline(xintercept = 2024.5,
               linetype = "dotted", color = "grey40", linewidth = 0.6) +
    annotate("text", x = 2024.6, y = Inf,
             label = "Forecast \u2192",
             hjust = 0, vjust = 1.8, size = 3, color = "grey40") +
    scale_color_manual(values = colors, name = "Month") +
    scale_linetype_manual(
      values = c("Historical" = "solid", "Forecast" = "dashed"),
      name   = NULL
    ) +
    scale_x_continuous(breaks = c(2015, 2018, 2021, 2024, 2027, 2030)) +
    labs(
      title    = paste(res$name,
                       "\u2014 Water Table Forecast: Selected Months"),
      subtitle = paste0(
        "Training: 2015\u20132024  |  ",
        "Solid: historical  |  Dashed: forecast (2025\u20132030)"
      ),
      x = "Year", y = "Water Table Depth (m)",
      caption = paste0(
        "FTSM (RAPCA, order = ", res$order, ", ARFIMA)  |  ",
        "Variance explained \u2014 ",
        paste0("PC", seq_along(res$varprop), ": ",
               round(res$varprop * 100, 1), "%",
               collapse = ",  ")
      )
    ) +
    theme_classic() +
    theme(
      plot.title      = element_text(size = 12, face = "bold"),
      plot.subtitle   = element_text(size = 9,  color = "grey30"),
      plot.caption    = element_text(size = 7.5, color = "grey40"),
      legend.position = "right",
      axis.text.x     = element_text(angle = 45, hjust = 1)
    )
}
```

## ----------------------------------------------------------
## Print all plots
## ----------------------------------------------------------

```{r}
for (i in seq_along(full_forecasts)) {
  print(plot_faceted(full_forecasts[[i]], all_plot_data[[i]]))
  print(plot_selected(full_forecasts[[i]], all_plot_data[[i]]))
}
```

## ----------------------------------------------------------
## Forecast value table (for appendix or supplementary)
## Rows = months, columns = forecast years, one table per district
## ----------------------------------------------------------

```{r}
cat("\n========== FORECAST VALUES (mean water table depth, m) ==========\n")
for (res in full_forecasts) {
  cat("\nDistrict:", res$name, "\n")
  tbl <- as.data.frame(round(res$fc_mat, 3))
  rownames(tbl) <- month_labels
  print(tbl)
}
```

